---
title: "Differential Expression Analysis - Impact of DSS and NFKB perturbation - Acute colitis"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: united
params:
  workdir: '/data/local/buyar/collaborations/marina_acute_colitis'
author: Bora Uyar
---

`r date()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, fig.width = 10,
                      fig.height = 8)
library(data.table)
library(ggplot2)
library(ggpubr)
library(pbapply)
library(RUVSeq)
library(EDASeq)
library(DESeq2)
library(progeny)
library(ggrepel)
library(pheatmap)
library(openxlsx)

ggplot2::theme_set(ggpubr::theme_pubclean())

workdir <- '/data/local/buyar/collaborations/marina_acute_colitis'
source(file.path(workdir, '/src/utils.R'))
sample_sheet <- read.csv(file.path(workdir, 'data/sample_sheet.csv'))
counts <- read.table(file.path(workdir, '/output/feature_counts/raw_counts/hisat2/counts.tsv'), header = T, check.names = F)
deseq_norm_counts <- read.table(file.path(workdir, 
                       '/output/feature_counts/normalized/hisat2/deseq_normalized_counts.tsv'), check.names = F)
colData <- data.frame(sample_sheet[,-1], row.names = sample_sheet$name)
colData$DSS <- do.call(c, lapply(strsplit(colData$sample_type, "_"), function(x) x[1]))
colData$NFKB <- do.call(c, lapply(strsplit(colData$sample_type, "_"), function(x) x[length(x)]))

# get id mappings
ids <- readRDS(file.path(workdir, 'data/ensmusg2name.RDS'))
```

# Before correction {.tabset}

Using DESeq2 normalized counts (before accounting for additional sources of variation)

It looks like primary impact is from DSS treatment, secondary impact is NFKB status (but weak)

## nfkb status (repressed or wt)

```{r}
P <- get_pca(log(deseq_norm_counts+1), colData)
ggplot(P$dat, aes(x = PC1, y = PC2)) + geom_point(aes(color = NFKB), size = 5) +
  geom_text_repel(aes(label = sample_type)) + 
  labs(x = P$var_exp[['PC1']], y = P$var_exp[['PC2']])
```

## DSS treatment (short / long / untreated)

```{r}
ggplot(P$dat, aes(x = PC1, y = PC2)) + geom_point(aes(color = DSS), size = 5) +
  geom_text_repel(aes(label = sample_type)) + 
  labs(x = P$var_exp[['PC1']], y = P$var_exp[['PC2']])
```

## NFKB status and DSS treatment

```{r}
ggplot(P$dat, aes(x = PC1, y = PC2)) + geom_point(aes(color = DSS, shape = NFKB), size = 5) +
  geom_text_repel(aes(label = sample_type)) + 
  labs(x = P$var_exp[['PC1']], y = P$var_exp[['PC2']])
```


# Detect unwanted variation and correct for those

```{r}
# use RUVs 
k <- 5
set <- newSeqExpressionSet(counts = as.matrix(counts),
                           phenoData = colData)
differences <- makeGroups(colData$sample_type)
set_s <- RUVs(set, unique(rownames(set)), 
              k=k, differences) #all genes

norm_counts <- log(normCounts(set_s)+1)


```

## Plot normalized counts {.tabset}

### by DSS

```{r}
P <- get_pca(norm_counts, colData)
ggplot(P$dat, aes(x = PC1, y = PC2)) + geom_point(aes(color = DSS), size = 5) +
  geom_text_repel(aes(label = sample_type)) + 
  labs(x = P$var_exp[['PC1']], y = P$var_exp[['PC2']])
```

### by NFKB

```{r}
ggplot(P$dat, aes(x = PC1, y = PC2)) + geom_point(aes(color = NFKB), size = 5) +
  geom_text_repel(aes(label = sample_type)) + 
  labs(x = P$var_exp[['PC1']], y = P$var_exp[['PC2']])

```

### by DSS & NFKB
```{r}
ggplot(P$dat, aes(x = PC1, y = PC2)) + geom_point(aes(color = DSS, shape = NFKB), size = 5) +
  geom_text_repel(aes(label = sample_type)) + 
  labs(x = P$var_exp[['PC1']], y = P$var_exp[['PC2']])
```

Now, I want to integrate the learned confounding variables to update DESeq2 analysis. 

```{r}
pdat <- pData(set_s)
pdat <- pdat[,c('sample_type', 'NFKB', 'DSS', grep('W_', colnames(pdat), value = T))]
pdat$group <- as.factor(pdat$sample_type)
colnames(pdat) <- gsub("_", "", colnames(pdat))
myFormula <- as.formula(paste0(" ~ group + ", paste(grep('^W', colnames(pdat), value = T)[1:k], collapse = " + ")))
dds <- DESeqDataSetFromMatrix(countData = counts, 
                              colData = pdat, 
                              design = myFormula)
# remove uninformative features
dds <- dds[ rowSums(DESeq2::counts(dds)) > 10, ]
dds <- DESeq(dds)
```

Some more diagnostic plots from DESeq2 after incorporating additional factors for unwanted variation. 

```{r}
# diagnostics 
plotMA(dds)
plotDispEsts(dds)
```

For each pair of terms, do a DE analysis 

```{r}
#CNT: character vector of 3 elements: 1. variable of interest, 2. case factor 3. control factor
get_de <- function(dds, CNT, ids) {
  res <- results(dds, contrast = CNT)
  res <- res[order(res$padj),]
  res$geneName <- ids[match(rownames(res), id)]$name
  res <- list('dt' = data.table(data.frame(res), keep.rownames = T))
  names(res) <- paste0(CNT[2], "_vs_", CNT[3])
  return(res)
}

# do a DE analysis based on learned W factors 
comparisons <- data.table(expand.grid('group', unique(pdat$group), unique(pdat$group)))
comparisons <- comparisons[Var2 != Var3]
comparisons$index <- apply(comparisons, 1, function(x) paste0(sort(x[2:3]), collapse = ' '))
comparisons <- comparisons[,.SD[which.min(Var2)],by = index][order(Var3, decreasing = T)][,2:4]

results <- do.call(c, pbapply::pbapply(comparisons, 1, function(x) {
  get_de(dds, x, ids)
}))

# shorten the comparison names to fit 22 characters per sheet name
shorten <- function(x) {
  name <- gsub("_dss|eated|", "", x)
  name <- gsub("_wt", "WT", name)
  name <- gsub("_loxp", "LOXP", name)
  name <- gsub("_vs_", "_", name)
  return(name)
}



# save DE tables to excel file
OUT1 <- createWorkbook() # for complete tables
OUT2 <- createWorkbook() # for only significant results
dummy <- lapply(names(results), function(comp) {
  # shorten the sheet names
  sname <- shorten(comp)
  addWorksheet(OUT1, sname)
  writeData(OUT1, sheet = sname, x = results[[comp]])
  addWorksheet(OUT2, sname)
  writeData(OUT2, sheet = sname, x = results[[comp]][padj < 0.01])
})
saveWorkbook(OUT1, file.path(workdir, "DE_results.xlsx"), overwrite = T)
saveWorkbook(OUT2, file.path(workdir, "DE_results.significant_genes.xlsx"), overwrite = T)


```

Plot number of DE genes in each comparison
```{r}
# count # of DE genes per comparison
DEstats <- do.call(rbind, lapply(names(results), function(x) {
  case <- unlist(strsplit(x, "_vs_"))[1]
  ctrl <- unlist(strsplit(x, "_vs_"))[2]
  data.table('case' = case, 'ctrl' = ctrl, 'DEgenes' = nrow(results[[x]][padj < 0.01]))
}))

ggplot(DEstats, aes(x = ctrl, y = case)) + 
  geom_tile(aes(fill = DEgenes)) + 
  geom_text(aes(label = DEgenes), color = 'white', size = 8)

```

# Genes of interest 

We have a list of genes with expected changes (from qPCR) and a set of house-keeping genes. 
Make a plot of the DE stats of these genes:
```{r}
goi <- list('house_keeping' = c('Hprt', 'Tbp'),
            'expected_to_change' = c('Tnfaip3', 'Nfkbia', 'Tnf'))

dt <- do.call(rbind, lapply(names(results), function(x) {
  dt <- results[[x]][geneName %in% paste(unlist(goi))]
  dt$comp <- x
  dt$type <- ifelse(dt$geneName %in% goi$house_keeping, "houseKeeping", "others")
  return(dt)
}))

ggplot(dt, aes(x = comp, y = -log10(padj))) + 
  geom_bar(aes(fill = type), stat = 'identity') + 
  geom_hline(yintercept = -log10(0.01)) + 
  facet_wrap( ~ geneName, ncol = 5) + coord_flip()

```

# Plot expression distributions of  selected genes {.tabset}

```{r}
get_exp <- function(g) {
  gid <- ids[match(g, name)]$id
  exp <- data.table('normalized_exp' = norm_counts[gid,], 'group' = colData[colnames(norm_counts),]$sample_type)
  return(exp)
}
dt <- do.call(rbind, lapply(c('Tnf', 'Tnfaip3', 'Hprt', 'Tbp', 'Nfkbia'), function(g) {
  dt <- get_exp(g)
  dt$gene <- g
  return(dt)
}))

ggboxplot(dt, x = 'group', y = 'normalized_exp', add = 'jitter', color = 'group', facet.by = 'gene', scales = 'free') +
  coord_flip()
```


# Gene-set scores

We can also compute cell type signature scores in each sample and see their clustering

```{r}
# gene set scoring - use cell type signatures from msigdb for mouse
gs <- read_msigdb('/data/local/buyar/collaborations/marina_kol/data/genesets/m8.all.v2022.1.Mm.symbols.gmt')
#gs <- read_msigdb('/data/local/buyar/collaborations/marina_kol/data/genesets/m2.all.v2022.1.Mm.symbols.gmt')
#gs <- read_msigdb('/data/local/buyar/collaborations/marina_kol/data/genesets/m3.gtrd.v2022.1.Mm.symbols.gmt')
# convert gene sets from gene names to ens ids
gs_ids <- sapply(gs, function(x) {
  na.omit(unique(ids[match(x, name)]$id))
})
# rank genes (based on ruvs normalized counts)
norm_counts <- log(normCounts(set_s)+1)
rankData <- singscore::rankGenes(expreMatrix = norm_counts)
# score gene sets 
scores <- pbapply::pbsapply(gs_ids, function(x) {
  score_gene_set(rankData, genes_up = x)
})
rownames(scores) <- colnames(rankData)

# visualize scores
pheatmap(scores, annotation_row = pdat[rownames(scores), 'group',drop=F], 
         show_colnames = F, scale = 'column')
# also print to file
pheatmap(t(scores), annotation_col = pdat[rownames(scores), 'group',drop=F], scale = 'row', 
         fontsize_row = 5, fontsize_col = 6, 
         width = 12, height = 14, border_color = NA, cutree_cols = 6, cutree_rows = 6,
         filename = 'msigdb_celltype_signatures.scores.pdf', )

```

See which gene sets show distinction across groups

```{r}
res <- compute_anova(scores, pdat[rownames(scores),'group'])
dt <- data.table(scores[,res[order(padj)][1:6]$variable], 
                 'group' = pdat[rownames(scores), 'group'])
mdt <- melt.data.table(dt, id.vars = 'group')
mdt$variable <- gsub("TABULA_MURIS_SENIS", "TMS", mdt$variable)

ggboxplot(mdt, x = 'group', y = 'value', add = 'jitter', color = 'group') + 
  facet_wrap(~ variable, scales = 'free') + 
  coord_flip()
```

# Progeny - Perturbed Pathway Scores

For each DE analysis, we can compute which pathways are most likely to be perturbed compared to the control. 

```{r}
# run progeny 
# res: deseq result table
get_progeny_scores <- function(res) {
  # pick best result per gene name  
  # some ids are mapped to gene names ambigously 
  res <- res[,.SD[which.min(padj)],by = geneName][,c('geneName', 'stat')]
  m <- as.matrix(data.frame(res[,-1], row.names = res$geneName))
  progeny_scores <- data.frame(t(progeny(m,
                            scale=TRUE, organism="Mouse", 
                            top = 100, perm = 10000, z_scores = FALSE)))
  return(progeny_scores)
}

# from each DE result, compute progeny scores
progeny_scores <- pbapply::pbsapply(simplify = F, results, function(dt) {
 get_progeny_scores(dt) 
})

PS <- do.call(cbind, lapply(names(progeny_scores), function(x) {
  dt <- progeny_scores[[x]]
  colnames(dt)[1] <- x
  return(dt)
}))

pheatmap(t(PS))

```


```{r}
sessionInfo()
```



