---
title: "Manuscript Figures - Cell-Cell Communication Analysis using CellChat"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: readable
params:
  cellchat_file: ''
  
author: Bora Uyar
---

`r date()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, fig.height = 8, fig.width = 10)
library(CellChat)
library(ggplot2)
library(ggalluvial)
```

```{r}
# cell chat results per condition
cellchat <- readRDS(params$cellchat_file)

```

# Detected Cell-Cell Interactions 

## Interaction strength {.tabset}

### Untreated 
```{r}
groupSize <- as.numeric(table(cellchat$Untreated@idents))
netVisual_circle(cellchat$Untreated@net$weight, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, title.name = "Interaction weights/strength", 
                 vertex.label.cex = 1)
pdf(file = 'untreated.pdf', width = 10, height = 10)
netVisual_circle(cellchat$Untreated@net$weight, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, title.name = "Interaction weights/strength", 
                 vertex.label.cex = 3.2, edge.width.max = 15)
dev.off()
```


### DSS 
```{r}
groupSize <- as.numeric(table(cellchat$DSS@idents))
netVisual_circle(cellchat$DSS@net$weight, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, title.name = "Interaction weights/strength")

pdf(file = 'DSS.pdf', width = 10, height = 10)
netVisual_circle(cellchat$DSS@net$weight, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, title.name = "Interaction weights/strength",
                 vertex.label.cex = 3.2, edge.width.max = 15)
dev.off()

```

### Recovery
```{r}
groupSize <- as.numeric(table(cellchat$Recovery@idents))
pdf(file = 'recovery.pdf', width = 10, height = 10)
netVisual_circle(cellchat$Recovery@net$weight, vertex.weight = groupSize, 
                 weight.scale = T, label.edge= F, title.name = "Interaction weights/strength",
                 vertex.label.cex = 3.2, edge.width.max = 15)
dev.off()

```


## As a heatmap {.tabset}

### Untreated

```{r}
netVisual_heatmap(cellchat$Untreated, color.heatmap = "Reds", measure = 'weight')
```

### DSS

```{r}
netVisual_heatmap(cellchat$DSS, color.heatmap = "Reds", measure = 'weight')
```

### Recovery

```{r}
netVisual_heatmap(cellchat$Recovery, color.heatmap = "Reds", measure = 'weight')
```

# Pathways  

## L-R pair contributions {.tabset}

### Untreated
```{r, fig.height=10}
obj <- cellchat$Untreated
paths <- obj@netP$pathways
netAnalysis_contribution(obj, signaling = paths)
```

### DSS
```{r, fig.height=10}
obj <- cellchat$DSS
paths <- obj@netP$pathways
netAnalysis_contribution(obj, signaling = paths)
```

### Recovery
```{r, fig.height=10}
obj <- cellchat$Recovery
paths <- obj@netP$pathways
netAnalysis_contribution(obj, signaling = paths)
```


## Chord diagram {.tabset}

### Untreated 

```{r, fig.height=10}
obj <- cellchat$Untreated
netVisual_chord_gene(obj, slot.name = "netP", legend.pos.x = 10, lab.cex = 0.4)
```

### DSS

```{r, fig.height=10}
obj <- cellchat$DSS
netVisual_chord_gene(obj, slot.name = "netP", legend.pos.x = 10, lab.cex = 0.4)
```

### Recovery

```{r, fig.height=10}
obj <- cellchat$Recovery
netVisual_chord_gene(obj, slot.name = "netP", legend.pos.x = 10, lab.cex = 0.4)
```


```{r, echo=F}
# print to pdf 
pdf(file = 'chord_diagram.pdf', width = 20, height = 20)
netVisual_chord_gene(cellchat$Untreated, slot.name = "netP", legend.pos.x = 10, lab.cex = 1.5)
netVisual_chord_gene(cellchat$DSS, slot.name = "netP", legend.pos.x = 10, lab.cex = 1.2)
netVisual_chord_gene(cellchat$Recovery, slot.name = "netP", legend.pos.x = 10, lab.cex = 1)
dev.off()

# I use "epithelial cells" as the source and draw connections to all possible cell types 
# I put a thresolhd (reduce) to remove rare ones to avoid label overlaps 
pdf(file = 'chord_diagram_epithelial_as_source.pdf', width = 12, height = 10)
netVisual_chord_gene(cellchat$Untreated, sources.use = 3, slot.name = "netP", legend.pos.x = 10, lab.cex = 1.6)
netVisual_chord_gene(cellchat$DSS, sources.use = 3, slot.name = "netP", legend.pos.x = 10, lab.cex = 1.6, 
                     reduce = 0.005)
netVisual_chord_gene(cellchat$Recovery, sources.use = 3, slot.name = "netP", legend.pos.x = 10, lab.cex = 1.6,
                     reduce = 0.005)
dev.off()
```

# Network Analysis

```{r}
cellchat <- sapply(simplify = F, cellchat, function(obj) {
  netAnalysis_computeCentrality(obj, slot.name = "netP")
})
```

## Outgoing signaling patterns {.tabset}

### Untreated
```{r, fig.height=10}
obj <- cellchat$Untreated
netAnalysis_signalingRole_heatmap(obj, pattern = "outgoing", color.heatmap = 'Reds', 
                                  height = 15, width = 20)
```

### DSS
```{r, fig.height=10}
obj <- cellchat$DSS
netAnalysis_signalingRole_heatmap(obj, pattern = "outgoing", color.heatmap = 'Reds', 
                                  height = 15, width = 20)
```

### Recovery
```{r, fig.height=10}
obj <- cellchat$Recovery
netAnalysis_signalingRole_heatmap(obj, pattern = "outgoing", color.heatmap = 'Reds',
                                   height = 15, width = 20)
```


## Incoming signaling patterns {.tabset}

### Untreated
```{r}
obj <- cellchat$Untreated
netAnalysis_signalingRole_heatmap(obj, pattern = "incoming", color.heatmap = 'Reds',
                                   height = 15, width = 20)
```

### DSS
```{r}
obj <- cellchat$DSS
netAnalysis_signalingRole_heatmap(obj, pattern = "incoming", color.heatmap = 'Reds',
                                   height = 15, width = 20)
```

### Recovery
```{r}
obj <- cellchat$Recovery
netAnalysis_signalingRole_heatmap(obj, pattern = "incoming", color.heatmap = 'Reds',
                                   height = 15, width = 20)
```

## Incoming interaction patterns 
```{r}
nPatterns = 4
cellchat <- sapply(simplify = F, cellchat, function(obj) {
  identifyCommunicationPatterns(obj, pattern = "incoming", k = nPatterns, heatmap.show = F)
})
```

### River Plot {.tabset}

#### Untreated

```{r}
obj <- cellchat$Untreated
netAnalysis_river(obj, pattern = "incoming")
```

#### DSS 

```{r}
obj <- cellchat$DSS
netAnalysis_river(obj, pattern = "incoming")
```

#### Recovery

```{r}
obj <- cellchat$Recovery
netAnalysis_river(obj, pattern = "incoming")
```

## Outgoing interaction patterns

```{r}
nPatterns = 4
cellchat <- sapply(simplify = F, cellchat, function(obj) {
  identifyCommunicationPatterns(obj, pattern = "outgoing", k = nPatterns, heatmap.show = F)
})
```

### River Plot {.tabset}

#### Untreated

```{r}
obj <- cellchat$Untreated
netAnalysis_river(obj, pattern = "outgoing")
```

#### DSS 

```{r}
obj <- cellchat$DSS
netAnalysis_river(obj, pattern = "outgoing")
```

#### Recovery

```{r}
obj <- cellchat$Recovery
netAnalysis_river(obj, pattern = "outgoing")
```


```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
  var images = document.querySelectorAll('img');
  images.forEach(function(img, index) {
    var button = document.createElement('button');
    button.textContent = 'Download as PDF';
    button.style = 'display: block; margin-top: 10px;';

    button.onclick = function() {
      var jsPDF = document.createElement('script');
      jsPDF.src = 'https://cdnjs.cloudflare.com/ajax/libs/jspdf/2.3.1/jspdf.umd.min.js';
      jsPDF.onload = function() {
        var { jsPDF } = window.jspdf;
        const doc = new jsPDF({
          orientation: 'portrait',
          unit: 'pt',
          format: 'a4'
        });

        var imgEl = new Image();
        imgEl.onload = function() {
          // Calculate scale ratio to fit the image within the page
          var scaleWidth = 595 / this.width;
          var scaleHeight = 842 / this.height;
          var scale = Math.min(scaleWidth, scaleHeight);

          // Calculate dimensions to maintain aspect ratio
          var imgWidth = this.width * scale;
          var imgHeight = this.height * scale;

          // Position the image centered horizontally and vertically
          var xPos = (595 - imgWidth) / 2;
          var yPos = (842 - imgHeight) / 2;

          doc.addImage(this, 'PNG', xPos, yPos, imgWidth, imgHeight);
          doc.save('download-' + index + '.pdf');
        };
        imgEl.crossOrigin = "Anonymous";
        imgEl.src = img.src;
      };
      document.body.appendChild(jsPDF);
    };

    img.parentNode.insertBefore(button, img.nextSibling);
  });
});
```