---
title: "Analysis of T-regs - Single-cell Data - 3rd Batch"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: united
params:
  seu_file: ''

author: Bora Uyar
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, fig.width = 10,
                      fig.height = 8)
library(data.table)
library(Seurat)
library(openxlsx)
library(ggplot2)

ggplot2::theme_set(ggpubr::theme_pubclean())

seu <- readRDS(params$seu_file)
```


# All Cells {.tabset}

## by Sample

```{r}
DimPlot(seu, group.by = 'sample')
```

## by Cell Type 

```{r}
DimPlot(seu, group.by = 'celltype', label = T, label.size = 6)
```

## by Cell Type vs Sample
```{r}
DimPlot(seu, group.by = 'celltype', split.by = 'sample', label = T, label.size = 4, ncol = 2)
```

# Regulatory T-Cells 

```{r}
seu_tregs <- seu[,which(seu$celltype == 'Tregs')]
```

## Foxp3 

Tregs were selected based on Foxp3 expression in the clusters that overlap T-cells. 

```{r}
FeaturePlot(seu, "Foxp3")
```

## Marker analysis 

We carry out differential marker analysis between: 

1. KO_COLON vs WT_COLON
2. KO_SPLEEN vs WT_SPLEEN

```{r}
# marker discovery within Treg subpopulation
colon <- data.table(FindMarkers(seu_tregs, 
                             min.pct = 0.1, 
                             logFoldChange = 0.25,
                             test.use = 'wilcox', 
                             ident.1 = 'KO_COLON',
                             ident.2 = 'WT_COLON',
                             group.by = 'sample'), 
                 keep.rownames = T)[order(p_val)]

spleen <- data.table(FindMarkers(seu_tregs, 
                                   min.pct = 0.1, 
                                   logFoldChange = 0.25,
                                   test.use = 'wilcox', 
                                   ident.1 = 'KO_SPLEEN',
                                   ident.2 = 'WT_SPLEEN',
                                   group.by = 'sample'), 
                       keep.rownames = T)[order(p_val)]
```

### Colon markers (KO vs WT) in Tregs {.tabset}

We find markers in colon samples (KO vs WT) and look at their patterns across samples in T-reg cells. 

#### Top Markers (padj < 0.1)

```{r}
colon_genes <- colon[p_val_adj < 0.1]$rn
knitr::kable(colon[p_val_adj < 0.1], caption = 'Colon Marker Genes KO vs WT')
```

#### DotPlot
```{r}
DotPlot(seu_tregs, colon_genes, group.by = 'sample') + coord_flip()
```

#### Heatmap 

```{r}
DoHeatmap(seu_tregs, colon_genes, group.by = 'sample')
```

#### VlnPlot

```{r}
VlnPlot(seu_tregs, colon_genes, group.by = 'sample', ncol = 4)
```


### Spleen markers (KO vs WT) in Tregs {.tabset}

We find markers in colon samples (KO vs WT) and look at their patterns across samples in T-reg cells. 

#### Top Markers (padj < 0.1)

```{r}
spleen_genes <- spleen[p_val_adj < 0.1]$rn
knitr::kable(spleen[p_val_adj < 0.1],  caption = 'Spleen Marker Genes KO vs WT')
```

#### DotPlot
```{r}
DotPlot(seu_tregs, spleen_genes, group.by = 'sample') + coord_flip()
```

#### Heatmap 

```{r}
DoHeatmap(seu_tregs, spleen_genes, group.by = 'sample')
```

#### VlnPlot (only showing top 10)

```{r}
VlnPlot(seu_tregs, spleen_genes[1:10], group.by = 'sample', ncol = 5)
```


### Common Markers in Both Spleen & Colon T-regs

We find Ftl1 and Gdpd3 as common markers with a clear upregulation in T-reg cells of KO samples in both colon and spleen tissues compared to WT.  

```{r}
common <- intersect(colon_genes, spleen_genes)
```

#### Feature Plot
```{r}
FeaturePlot(seu_tregs, common, split.by = 'sample')
```

#### VlnPlot
```{r}
VlnPlot(seu_tregs, common, group.by = 'sample')
```



```{r save_tables}
# save marker results as xlsx tables 
# save DE tables to excel file
OUT <- createWorkbook() # for complete tables
sname <- "Colon_Markers_KOvsWT"
addWorksheet(OUT, sname)
writeData(OUT, sheet = sname, x = colon)
sname <- "Spleen_Markers_KOvsWT"
addWorksheet(OUT, sname)
writeData(OUT, sheet = sname, x = spleen)
saveWorkbook(OUT, file.path("Tregs_Markers.xlsx"), overwrite = T)
```


# Session Info

```{r}
print(sessionInfo())
```


```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
  var images = document.querySelectorAll('img');
  images.forEach(function(img, index) {
    var button = document.createElement('button');
    button.textContent = 'Download as PDF';
    button.style = 'display: block; margin-top: 10px;';

    button.onclick = function() {
      var jsPDF = document.createElement('script');
      jsPDF.src = 'https://cdnjs.cloudflare.com/ajax/libs/jspdf/2.3.1/jspdf.umd.min.js';
      jsPDF.onload = function() {
        var { jsPDF } = window.jspdf;
        const doc = new jsPDF({
          orientation: 'portrait',
          unit: 'pt',
          format: 'a4'
        });

        var imgEl = new Image();
        imgEl.onload = function() {
          // Calculate scale ratio to fit the image within the page
          var scaleWidth = 595 / this.width;
          var scaleHeight = 842 / this.height;
          var scale = Math.min(scaleWidth, scaleHeight);

          // Calculate dimensions to maintain aspect ratio
          var imgWidth = this.width * scale;
          var imgHeight = this.height * scale;

          // Position the image centered horizontally and vertically
          var xPos = (595 - imgWidth) / 2;
          var yPos = (842 - imgHeight) / 2;

          doc.addImage(this, 'PNG', xPos, yPos, imgWidth, imgHeight);
          doc.save('download-' + index + '.pdf');
        };
        imgEl.crossOrigin = "Anonymous";
        imgEl.src = img.src;
      };
      document.body.appendChild(jsPDF);
    };

    img.parentNode.insertBefore(button, img.nextSibling);
  });
});
```

